Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-28 paper.tex
Updates on parsimony algs

\subsection{Tree inference from clusters of samples}
For a given daily batch of sample sequences that are to be attached to the
ARG, the LS HMM provides the most likely copying path and required mutations
for each of those sequences. We could directly attach these samples to
the ARG according to the copying paths and sets of mutations, but
such an approach would be highly unparsimonious because there are
often many samples that share the same copying path, with
complex patterns of shared mutations among those clusters of samples.
In this section we discuss the strategy used to infer likely tree
topologies within these clusters of samples sharing the same copying path.
In the majority of cases the copying path will consist of a single
node (i.e., no recombination), and so for simplicity we will assume in the
following discussion that all of the samples in a HMM-defined cluster
attach to a single node. The special treatment afforded to
newly-detected putative recombinants are discussed below
in Section XXX.

@@

\subsection{Parsimony updates}
[Discuss the parsimony tweaks we make to the trees. Get a better name
for the subsection title]

>>>>

\subsection{Tree inference from HMM sample clusters}
For a given daily batch of sample sequences that are to be attached to the
ARG, the LS HMM provides the most likely copying path and required mutations
for each of those sequences. We could directly attach these samples to
the ARG according to the copying paths and sets of mutations, but
such an approach would be highly unparsimonious because there are
often many samples that share the same copying path, with
complex patterns of shared mutations among those clusters of samples.
In this section we discuss the strategy used to infer likely tree
topologies within these clusters of samples sharing the same copying path.
In the majority of cases the copying path will consist of a single
node (i.e., no recombination), and so for simplicity we will assume in the
following discussion that all of the samples in a HMM-defined cluster
attach to a single node. The special treatment afforded to
newly-detected putative recombinants are discussed below
in Section XXX.

@@

\subsection{Parsimony increasing hueristics}
Attaching trees built from the clusters of samples that copy from
a particular node (or path of nodes for recombinants, see the
next section) under the HMM is an inherently greedy strategy,
and can produce inferences that are clearly unparsimonious.
The final step in adding a daily batch of samples to the ARG
is then to perform some local updates that target specific
types of parsimony violation in the just-updated regions of the
ARG. There are currently two parsimony-increasing operations
applied, which we refer to as ``mutation collapsing'' and ``reversion
pushing''.

Mutation collapsing inspects the siblings of a newly attached node
to check if any of them share (a subset) of the mutations that
it carries. If so, we increase the overall parsimony of the inference
by creating a new node representing the ancestor that carried
those shared mutations, and make that new node the parent of the
siblings carrying those shared mutations. The patterns of
shared mutations between siblings can be complex, and the current
implementation uses a simple greedy strategy for choosing
the particular mutations to collapse. [TODO back ref to the
Results section to discuss how important these nodes are]

The reversion push operation inspects newly added nodes to see
if any of its mutations are ``immediate reversions''; that is,
are reversions of a mutation that occured on the new node's
immediate parent. We increase the overall parsimony of the
inference by ``pushing in'' a new node in which descends from the
original parent which carries all its mutations except those
causing the reversions on the newly added node.
[TODO back ref to the
Results section to discuss how important these nodes are]

Together these operations substantially increase the overall parsimony
of the inferred ARGs, most noticably in terms of the number of
reversion mutations inferred. However, they are both simple
greedy operations, just examining the local parts of the
ARG topology affected by newly added samples. Since the inferred
ARGs still contain a large number of reversion mutations which
are likely to be mostly artefactual, it is clear that there
is room for improvement, and that further parsimony increasing
hueristics would likely be of benefit.
